Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures

### 2021-04-28 doc/temp-sentiment.tex

\section{Background}
\subsection{Temperature and Mental Health}
Higher temperatures are associated with many indicators of worsened mental health.  Multiple studies have found strong evidence that higher temperatures are associated with increases in suicides in the United States \citep{Burke2018Aug, Mullins2019Dec, Dixon2007May}, and many others have demonstrated the same relationship around the world \citep{Qi2014Dec, Page2006Aug, Likhvar2011Jan}.  In addition to suicide, other research has found effects temperature on hospitalization events related to mental health disorders such as bipolar disorder and schizophrenia \citep{Mullins2019Dec, Lee2007Jan, Shapira2004Feb, Sung2013Feb, Gupta1992Jun, Hansen2008Oct}.

@@

% The pathways
Higher temperature associated with climate change could influence mental health directly by exposing people to trauma. It could also indirectly influence mental health by affecting (1) physical health and (2)community wellbeing\citep{RN1314}. For the direct influence pathway, extreme heat events or increasing temperatures have been associated with the increase in aggression,higher suicide rates and other hospital admissions(Anderson and Anderson 1998,Cheatwood 1995; Cohn et al. 2004,Bouchama et al.2007.). Heat exposures in working environments could also reduce people's capacity to deal with physical and mental task, and increase the risk of accidents, because of the excessive core body temperature and dehydration((Bridger 2003; Kerslake 1972; Kjellstrom et al.2009d,Leithead and Lind 1964,Schrier et al. 1970). The loss of work capacity would in turn result in loss of income, which could also cause mental health problems(Kjellstrom 2009b).

As for the indirect pathway, first, because of the reciprocal causation relationship between physical health problems and mental health problems(Miller et al. 2009; Prince et al. 2007), heat exposures associated with climate change along with other climate events and indirect health risks threatening physical health will directly influence mental health((Berry 2007,World Health Organization 2008b).

@@

% Talk more about race

@@

\section{Results}
We examined the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an indicator of heat stress that accounts for temperature, humidity, wind speed, and solar radiation.  We then explored how the racial and income characteristics of neighborhoods moderate the relationship between WBGT and sentiment.  Finally, we show how twitter sentiment is an important overall indicator of mental health by assessing the relationship between daily sentiment at the county level and the prevalence of crime and suicide.

@@

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../res/wbgt-income.png}
  \caption{Effect of changes in wet-bulb globe temperature on expressed sentiment relationship as moderated by neighborhood income.}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../res/wbgt-race_q.png}
  \caption{Effect of changes in wet-bulb globe temperature on expressed entiment relationship as moderated by neighborhood majority race.}
  \label{fig:sub2}
\end{subfigure}
\caption{How the relationship between WBGT and sentiment is moderated }
\label{fig:test}
\end{figure}

@@

\section{Data}
A proposed list of tables and figures:
Figures for location of USA Tweets (to show aggregation and coverage).
Figures of income blocks and temperature extremes.
Table 1. Variable descriptions for what's included in our models.
Table 2. Descriptive statistics for datasets/variables we include.
Table 3. Model results with variables included in the models coefficients, SEs, p-values, sample size.
Appendix table. Sentiment scores across all types we considered.
\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{../res/Boston_Map.png}
  \caption{Here is a map of all tweets in boston, overlaid with income by census block}
  \label{fig:timeseries}
\end{figure}

>>>>

\section{Background}
\subsection{Temperature and Mental Health}
%https://www.pnas.org/content/115/43/10953
Higher temperatures are associated with many indicators of worsened mental health.  Multiple studies have found strong evidence that higher temperatures are associated with increases in suicides in the United States \citep{Burke2018Aug, Mullins2019Dec, Dixon2007May}, and many others have demonstrated the same relationship around the world \citep{Qi2014Dec, Page2006Aug, Likhvar2011Jan}.  In addition to suicide, other research has found effects temperature on hospitalization events related to mental health disorders such as bipolar disorder and schizophrenia \citep{Mullins2019Dec, Lee2007Jan, Shapira2004Feb, Sung2013Feb, Gupta1992Jun, Hansen2008Oct}.

@@

% The pathways
Higher temperature associated with climate change could influence mental health directly by exposing people to trauma. It could also indirectly influence mental health by affecting (1) physical health and (2)community wellbeing\citep{RN1314}. For the direct influence pathway, extreme heat events or increasing temperatures have been associated with the increase in aggression,higher suicide rates and other hospital admissions(Anderson and Anderson 1998,Cheatwood 1995; Cohn et al. 2004,Bouchama et al.2007.). Heat exposures in working environments could also reduce people's capacity to deal with physical and mental task, and increase the risk of accidents, because of the excessive core body temperature and dehydration((Bridger 2003; Kerslake 1972; Kjellstrom et al.2009d,Leithead and Lind 1964,Schrier et al. 1970). The loss of work capacity would in turn result in loss of income, which could also cause mental health problems(Kjellstrom 2009b).%% more physiological terms

As for the indirect pathway, first, because of the reciprocal causation relationship between physical health problems and mental health problems(Miller et al. 2009; Prince et al. 2007), heat exposures associated with climate change along with other climate events and indirect health risks threatening physical health will directly influence mental health((Berry 2007,World Health Organization 2008b). Second，disordered temperature associated with climate change may destroy the economic in agricultural-production-dependent communities. For example, extreme heat reduces the work capacity of laborers in farm fields(Kjellstrom et al. 2009d), which further destroys agricultural-supported industries in local area(Berry et al. 2008). The following economic pressures would undermine social capital and then lead to mental health problems.

@@

% Talk more about race
% This is interesting: https://www.pnas.org/content/118/17/e2019624118.abstract?etoc

@@

\section{Results}

\subsection{Overall Effect}
We examined the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an indicator of heat stress that accounts for temperature, humidity, wind speed, and solar radiation.  We then explored how the racial and income characteristics of neighborhoods moderate the relationship between WBGT and sentiment.  Finally, we show how twitter sentiment is an important overall indicator of mental health by assessing the relationship between daily sentiment at the county level and the prevalence of crime and suicide.

@@

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../res/wbgt-income.png}
  \caption{Effect of changes in wet-bulb globe temperature on expressed sentiment relationship as moderated by neighborhood income.}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../res/wbgt-race_q.png}
  \caption{Effect of changes in wet-bulb globe temperature on expressed sentiment relationship as moderated by neighborhood majority race.}
  \label{fig:sub2}
\end{subfigure}
\caption{How the relationship between WBGT and sentiment is moderated }
\label{fig:test}
\end{figure}

\subsection{Effects by Combined Statistical Area}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.75\linewidth]{../res/map_wbgt.png}
  \caption{Effects of rising temperatures on sentiment by Combined Statistical Areas (CSAs) with over 1 million people.  The value shown is the predicted change in vader score for a 25-degree increase in temperatures.  CSAs with a statistically significant effect ($p < 0.1$) are outlined in black.}
  \label{fig:map-wbgt}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.75\linewidth]{../res/map_wbgt_income.png}
  \caption{Expected gap in sentiment scores between poor and rich neighborhoods, by Combined Statistical Areas (CSAs) with over 1 million people.  The value shown is the predicted gap in vader scores for a 25-degree increase in temperatures between neighborhoods at the 5th percentile and the 90th percentile for income.  CSAs with a statistically significant effect ($p < 0.1$) are outlined in black.}
  \label{fig:map-wbgt-income}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.75\linewidth]{../res/map_wbgt_race.png}
  \caption{Expected gap in sentiment scores between largely minority and largely white neighborhoods, by Combined Statistical Areas (CSAs) with over 1 million people.  The value shown is the predicted gap in vader scores for a 25-degree increase in temperatures between neighborhoods at the 5th percentile and the 95th percentile for whiteness.  CSAs with a statistically significant effect ($p < 0.1$) are outlined in black.}
  \label{fig:map-wbgt-race}
\end{figure}

\subsection{Effects by Time of Day}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.75\linewidth]{../res/ts.png}
  \caption{Effects of rising temperatures on sentiment by hour of the day, with a 95\% confidence interval.  The value shown is the predicted change in vader score for a 25-degree increase in temperatures.}
  \label{fig:map-wbgt}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.75\linewidth]{../res/ts_income.png}
  \caption{Expected gap in sentiment scores between largely minority and largely white neighborhoods, by hour of the day, with a 95\% confidence interval.  The value shown is the predicted gap in vader scores for a 25-degree increase in temperatures between neighborhoods at the 5th percentile and the 95th percentile for income.}
  \label{fig:map-wbgt-income}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.75\linewidth]{../res/ts_race.png}
  \caption{Expected gap in sentiment scores between largely minority and largely white neighborhoods, by hour of the day, with a 95\% confidence interval.  The value shown is the predicted gap in vader scores for a 25-degree increase in temperatures between neighborhoods at the 5th percentile and the 95th percentile for whiteness.}
  \label{fig:map-wbgt-race}
\end{figure}

@@

\section{Data}
A proposed list of tables and figures:
Figures for location of USA Tweets (to show aggregation and coverage).
Figures of income blocks and temperature extremes.
Table 1. Variable descriptions for what's included in our models.
Table 2. Descriptive statistics for datasets/variables we include.
Table 3. Model results with variables included in the models coefficients, SEs, p-values, sample size.
Appendix table. Sentiment scores across all types we considered.
